\documentclass{manualjp}

\title{Macintoshモジュールリファレンス}

\input{boilerplatejp}

\makeindex              % tell \index to actually write the .idx file
\makemodindex           % ... and the module index as well.


\begin{document}

\maketitle

\ifhtml
\chapter*{序\label{front}}
\fi

\input{copyrightjp}

\begin{abstract}

\noindent
%% This library reference manual documents Python's extensions for the
%% Macintosh.  It should be used in conjunction with the
%% \citetitle[../lib/lib.html]{Python Library Reference}, which documents
%% the standard library and built-in types.

このライブラリリファレンスマニュアルでは、Macintosh用のPython拡張に関
して詳しく記述します。\citetitle[../lib/lib.html]{Python Library Reference}
も同時に参照しながら利用するようにしてください。標準ライブラリと組み込み型はそちら
に詳しく書かれています。

%% This manual assumes basic knowledge about the Python language.  For an
%% informal introduction to Python, see the
%% \citetitle[../tut/tut.html]{Python Tutorial}; the
%% \citetitle[../ref/ref.html]{Python Reference Manual} remains the
%% highest authority on syntactic and semantic questions.  Finally, the
%% manual entitled \citetitle[../ext/ext.html]{Extending and Embedding
%% the Python Interpreter} describes how to add new extensions to Python
%% and how to embed it in other applications.

このマニュアルを読むにあたっては、Python言語の基礎知識を持っていること
が必要です。Pythonの肩のこらない入門編が必要なら、
\citetitle[../tut/tut.html]{Python Tutorial}を読んでください。
\citetitle[../ref/ref.html]{Python Reference Manual}は、構文や意味論に
関する疑問を解決するため、それなりに分かった人が読むべきものです。最
後にひとつ。インタプリタの拡張と組み込み
\citetitle[../ext/ext.html]{Extending and Embedding the Python Interpreter}
という名のマニュアルでは、Pythonへ新たに拡張機能を追加す
る方法と、他のアプリケーションに組み込む方法について述べています。

\end{abstract}

\tableofcontents


\input{using.tex}                       % Using Python on the Macintosh


\chapter{MacPython モジュール \label{macpython-modules}}
% MacPython Modules

%% The following modules are only available on the Macintosh, and are
%% documented here:

このドキュメントで記述されている次のモジュールは、いずれもMacintoshで
のみ利用可能です。

\localmoduletable

\input{libmac}
\input{libmacfs}
\input{libmacic}
\input{libmacos}
\input{libmacostools}
\input{libmacui}
\input{libframework}
\input{libautogil}

\input{scripting}

\input{toolbox}                         % MacOS Toolbox Modules
\input{libcolorpicker}

\input{undoc}                           % Undocumented Modules

\appendix
\chapter{歴史とライセンス} % History and License
\input{license}

\chapter{日本語訳について}
\input{jptranslation.tex}

%
%  The ugly "%begin{latexonly}" pseudo-environments are really just to
%  keep LaTeX2HTML quiet during the \renewcommand{} macros; they're
%  not really valuable.
%

%begin{latexonly}
\renewcommand{\indexname}{Module Index}
%end{latexonly}
\input{modmac.ind}      % Module Index

%begin{latexonly}
\renewcommand{\indexname}{Index}
%end{latexonly}
\input{mac.ind}         % Index

\end{document}
